Bayesian RKHS-based methods in functional regression\support{Support information of the article.}

### 2022-03-07 paper/paper.tex
Some corrections

The most common linear regression model is the classical \(L^2\)-model, originally introduced in the first edition of the book by~\citet{ramsay2005functional} \incomment{fue realmente así?}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \beta_0 + \dotprod{X}{\beta} + \epsilon = \beta_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
\incomment{mejor poner el producto con el subíndice \(\langle \cdot , \cdot \rangle_2\)?}~where \(\beta_0\in \R\) and \(\epsilon\) is an independent error term with certain simplifying assumptions. In this case, the main parameter \(\beta=\beta(t)\) is a member of the infinite-dimensional space \(L^2[0, 1]\). Aside from the fact that functional (i.e. infinite-dimensional) optimization is a hard problem in and of itself \incomment{cita?}, there are a number of more subtle drawbacks that arise in this \(L^2\)-framework. For instance, \(L^2[0, 1]\) is a big space that contains many non-smooth or ill-behaved functions, but incidentally the model given by~\eqref{eq:l2-linear-model} does not include a finite-dimensional model based on a linear combination of projections of \(X\) as a particular case \citep[see][]{berrendero2019rkhs}. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

@@

@@

The value of \(p\), the dimensionality of the model, will be fixed beforehand in a suitable way (see Section~\ref{sec:model-choice} for details), and we will regard \(\beta_j\) and \(t_j\) as free parameters. Thus, we will search for our functional parameter in the set
\[
\mathcal H_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_1,\dots,\beta_j \in \R, \ t_1,\dots,t_j \in [0, 1]\right\}.
\]

@@

\begin{description}
  \item[Point estimates] If we consider a point-estimate statistic \(T\) to act as a summary of the posterior distribution, we can get the corresponding estimates \(\hat{\theta}=(\hat \beta, \hat \tau, \hat{\alpha}_0, \hat{\sigma}^2) = T(\{\theta^{(m)*}\})\) and predict the responses in the usual way by
  \[
    \hat Y_i = \hat{\alpha}_0 + \sum_{j=1}^p \hat{\beta}_j X_i(\hat{t}_j), \quad i=1,\dots, n'.
  \]
  The choice of a specific summary statistic results in a veiled assumption of an underlying loss function between the estimated and real parameters. In general, the mean is more sensitive to outliers while the median is more robust, but the latter assumes a \(L^1\)-type loss function. On the other hand, the mode is also a good candidate because it represents the point of highest probability. At any rate, these decisions are strongly dependent on several factors such as the skewness or the number of modes in the resulting posterior distribution, and thus should be made on a case-by-case basis.
  \item[Distributional estimates] We can also look at the approximate posterior distribution as a whole and compute the distribution \(Y^*\) of the ``sampled responses''
  \[
  \left\{Y_i^{(m)*} \equiv Y_i \mid X_i, \theta^{(m)*}:\ i=1,\dots,n', \ m=1,\dots,M\right\}
  \]
  following model~\eqref{eq:rkhs-model-linear-2}. Then, we can take the mean of all such samples as a proxy for the response:
  \[
  \hat Y_i = \frac{1}{M}\sum_{m=1}^M Y_i^{(m)*}, \quad i=1,\dots,n'.
  \]
  \item[Variable selection estimates] Lastly, we can focus only on the marginal posterior distribution of \(\tau\mid\mathcal D_n\) and perform variable selection via a point-estimate statistic \(T\). Specifically, we can consider the times \(\hat t_j = T(\{t_j^{(m)*}\})\) and reduce the original data set to the scalar variables \(\{X_i(\hat t_j):\ i=1, \dots,n, \ j=1,\dots,p\}\). After this selection, we can treat the problem as a finite-dimensional regression task and use any of the well-known multivariate algorithms suited for this situation.
\end{description}

@@

In the case of functional logistic regression, we can view the binary response variable \(Y\in\{0, 1\}\) as a Bernoulli random variable given the regressor \(X\), and consider an RKHS approach given by the following functional logistic equation:
\begin{equation}\label{eq:rkhs-model-logistic}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \Psi^{-1}_X(\alpha)\}}, \quad \alpha \in \mathcal H_{0,p}(K).
\end{equation}
Following our previous approach, after incorporating the sample information we can rewrite the classification model~\eqref{eq:rkhs-model-logistic} as
\begin{equation}\label{eq:rkhs-model-logistic-2}
Y_i \mid X_i,\theta \ \stackrel{\text{i.i.d.}}{\sim} \operatorname{Bernoulli}(p_i), \quad i=1,\dots, n,
\end{equation}
with
\begin{equation}\label{eq:rkhs-model-logistic-2-parameter}
  p_i = \mathbb P(Y_i=1 \mid X_i,\theta) = \frac{1}{\displaystyle 1 + \exp\left\{-\alpha_0 - \sum_{j=1}^p \beta_j X_i(t_j)\right\}}, \quad i=1,\dots, n,
\end{equation}
where in turn \(\beta_j\in\R\), \(t_j\in[0, 1]\) and \(\alpha_0\in\R\).

@@

@@

Another thing to consider is that in this case we are approximating a probability, so before producing an output we need to transform the predicted values to \(\{0, 1\}\). We will follow the standard convention of considering a threshold \(\gamma = 0.5\) and predicting \(\hat Y=1\) if and only if \(\mathbb P(Y=1\mid X) \geq \gamma\). With this thresholding strategy in mind, the point estimation approach is analogous to the linear regression case:
\[
\hat Y_i = \mathbb I \left( \left[\displaystyle 1 + \exp\left\{-\hat\alpha_0 - \sum_{j=1}^p \hat\beta_j X_i(\hat t_j)\right\}\right]^{-1} \geq \gamma \right), \quad i=1,\dots,n',
\]
where \(\mathbb I\) is the indicator function and the hat estimates are obtained through a summary statistic of the corresponding marginal posterior distributions. On the other hand, the distributional estimates are somewhat different now, since there is the question of which response (the Bernoulli variables in~\eqref{eq:rkhs-model-logistic-2} or the raw probabilities in~\eqref{eq:rkhs-model-logistic-2-parameter}) to consider when averaging the posterior samples:
\begin{description}
  \item[Average sampled probability] If we choose to average the sampled probabilities \(p_i^{(m)*} = \mathbb P(Y_i =1 \mid X_i,\theta^{(m)*})\), we get the following estimator:
  \[
    \hat Y_i = \mathbb I\left(\frac{1}{M} \sum_{m=1}^M p_i^{(m)*} \geq \gamma\right), \quad i=1,\dots,n'.
  \]
  \item[Sampled majority vote] If we choose to average the sampled binary responses \(Y^*\) instead, we compute the predictions as
\[
  \hat Y_i = \mathbb I\left(\frac{1}{M} \sum_{m=1}^M Y_i^{(m)*} \geq \gamma\right), \quad i=1,\dots,n',
\]
where each \(Y_i^{(m)*}\) follows a Bernoulli distribution of parameter \(p_i^{(m)*}\), for \(m=1,\dots,M\).
\end{description}
Lastly, the variable selection method is essentially the same as in the case of linear regression: we select \(p\) times from each trajectory based on the posterior distribution of \(\tau\mid \mathcal D_n\), and then feed the reduced dataset to a multivariate binary classification procedure.

@@

As for the default values of the rest of hyperparameters in~\eqref{eq:prior-linear}, several comments can be made:
\begin{itemize}
  \item For the expected value \(b_0\) we propose to use (an approximation of) the MLE of \(\beta\) (see Section~\ref{sec:results} for details on the implementation). Our numerical studies showed that the results were much better with this choice than, say, with a random or null vector.
  \item We found that the parameter \(g\) does not have as much influence on the final result, and the experimentation suggests that \(g=5\) is a good value.
  \item Lastly, we observed that the choice of \(\eta\) can have a considerable impact on the final estimator. That is why, in an effort to normalize its scale, we consider a compound parameter \(\eta = \tilde \eta \lambda_{\max}(G_\tau)\), where \(\lambda_{\max}(G_\tau)\) is the largest eigenvalue of the matrix \(G_\tau\) and \(\tilde\eta > 0\) is the actual tuning parameter. This standardization technique was used previously in the literature; see for example \citet{grollemund2019bayesian}.
\end{itemize}

>>>>

The most common functional linear regression model is the classical \(L^2\)-model, originally introduced in the first edition of the book by~\citet{ramsay2005functional} \incomment{fue realmente así?}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \beta_0 + \dotprod{X}{\beta} + \epsilon = \beta_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
\incomment{mejor poner el producto con el subíndice \(\langle \cdot , \cdot \rangle_2\)?}~where \(\beta_0\in \R\) and \(\epsilon\) is an independent error term with certain simplifying assumptions. In this case, the main parameter \(\beta=\beta(t)\) is a member of the infinite-dimensional space \(L^2[0, 1]\). Aside from the fact that functional (i.e. infinite-dimensional) optimization is a hard problem in and of itself \incomment{cita?}, there are a number of more subtle drawbacks that arise in this \(L^2\)-framework. For instance, \(L^2[0, 1]\) is a big space that contains many non-smooth or ill-behaved functions, but incidentally the model given by~\eqref{eq:l2-linear-model} does not include a finite-dimensional model based on a linear combination of projections of \(X\) as a particular case \citep[see][]{berrendero2019rkhs}. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

@@

\begin{comment}
  Quizás falta comentar algo de los resultados experimentales...
\end{comment}

@@

The value of \(p\), the dimensionality of the model, will be fixed beforehand in a suitable way (see Section~\ref{sec:model-choice} for details), and we will regard \(\beta_j\) and \(t_j\) as free parameters. Thus, we will search for our functional parameter in the set
\[
\mathcal H_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_1,\dots,\beta_p \in \R, \ t_1,\dots,t_p \in [0, 1]\right\}.
\]

@@

\begin{description}
  \item[Point estimates] If we consider a point-estimate statistic \(T\) to act as a summary of the posterior distribution, we can get the corresponding estimates \(\hat{\theta}=(\hat \beta, \hat \tau, \hat{\alpha}_0, \hat{\sigma}^2) = T(\{\theta^{(m)*}\})\) and predict the responses in the usual way by
  \[
    \hat Y_i = \hat{\alpha}_0 + \sum_{j=1}^p \hat{\beta}_j X_i(\hat{t}_j), \quad i=1,\dots, n'.
  \]
  The choice of a specific summary statistic results in a veiled assumption of an underlying loss function between the estimated and real parameters. In general, the mean is more sensitive to outliers while the median is more robust, but the latter assumes a \(L^1\)-type loss function. On the other hand, the mode is also a good candidate because it represents the point of highest probability. At any rate, these decisions are strongly dependent on several factors such as the skewness or the number of modes in the resulting posterior distribution, and thus should be made on a case-by-case basis.
  \item[Distributional estimates] We can also look at the approximate posterior distribution as a whole and compute the distribution of the ``sampled responses'' at each step of the chain following model~\eqref{eq:rkhs-model-linear-2}:
  \[
  Y^{(m)*} := \left\{Y_i^{(m)*} \equiv Y_i \mid X_i, \theta^{(m)*}:\ i=1,\dots,n'\right\}, \quad m=1,\dots,M.
  \]
  Then, we can take the mean of all such samples as a proxy for each response variable:
  \[
  \hat Y_i = \frac{1}{M}\sum_{m=1}^M Y_i^{(m)*}, \quad i=1,\dots,n'.
  \]
  \item[Variable selection estimates] Lastly, we can focus only on the marginal posterior distribution of \(\tau\mid\mathcal D_n\) and perform variable selection via a point-estimate statistic \(T\). Specifically, we can consider the times \(\hat t_j = T(\{t_j^{(m)*}\})\) and reduce the original data set to the scalar variables \(\{X_i(\hat t_j):\ i=1, \dots,n, \ j=1,\dots,p\}\). After this selection, we can treat the problem as a finite-dimensional regression task and use any of the well-known multivariate algorithms suited for this situation.
\end{description}

@@

In the case of functional logistic regression, we can view the binary response variable \(Y\in\{0, 1\}\) as a Bernoulli random variable given the regressor \(X=x \in L^2[0, 1]\), and suppose that \(\log\left(p(x)/(1-p(x))\right)\) is linear in \(x\), where \(p(x)=\mathbb P(Y=1\mid X=x)\). Then, an RKHS approach given by the following functional logistic equation:
\begin{equation}\label{eq:rkhs-model-logistic}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \Psi^{-1}_X(\alpha)\}}, \quad \alpha \in \mathcal H_{0,p}(K).
\end{equation}
Following our previous strategy, after incorporating the sample information we can rewrite the classification model~\eqref{eq:rkhs-model-logistic} as
\begin{equation}\label{eq:rkhs-model-logistic-2}
Y_i \mid X_i,\theta \ \stackrel{\text{i.i.d.}}{\sim} \operatorname{Bernoulli}(p_i), \quad i=1,\dots, n,
\end{equation}
with
\begin{equation}\label{eq:rkhs-model-logistic-2-parameter}
  p_i = \mathbb P(Y_i=1 \mid X_i,\theta) = \frac{1}{\displaystyle 1 + \exp\left\{-\alpha_0 - \sum_{j=1}^p \beta_j X_i(t_j)\right\}}, \quad i=1,\dots, n,
\end{equation}
where in turn \(\beta_j\in\R\), \(t_j\in[0, 1]\) and \(\alpha_0\in\R\).

@@

\begin{comment}
  The proposed model holds whenever \(X_0\), \(X_1\) are homoscedastic Gaussian processes (Bueno-Larraz et al.).
\end{comment}

@@

Another thing to consider is that in this case we are approximating a probability, so before producing an output we need to transform the predicted values to \(\{0, 1\}\). We will follow the standard convention of considering a threshold \(\gamma = 0.5\) and predicting \(\hat Y=1\) if and only if \(\mathbb P(Y=1\mid X) \geq \gamma\). With this thresholding strategy in mind, the point estimation approach is analogous to the linear regression case:
\[
\hat Y_i = \mathbb I \left( \left[\displaystyle 1 + \exp\left\{-\hat\alpha_0 - \sum_{j=1}^p \hat\beta_j X_i(\hat t_j)\right\}\right]^{-1} \geq \gamma \right), \quad i=1,\dots,n',
\]
where \(\mathbb I\) is the indicator function and the hat estimates are obtained through a summary statistic of the corresponding marginal posterior distributions. On the other hand, the distributional estimates are somewhat different now, since there is the question of which response (the Bernoulli variables in~\eqref{eq:rkhs-model-logistic-2} or the raw probabilities in~\eqref{eq:rkhs-model-logistic-2-parameter}) to consider when averaging the posterior samples:
\begin{description}
  \item[Average sampled probability] If we choose to average the sampled probabilities \(p_i^{(m)*} = \mathbb P(Y_i =1 \mid X_i,\theta^{(m)*})\), we get the following estimator:
  \[
    \hat Y_i = \mathbb I\left(\frac{1}{M} \sum_{m=1}^M p_i^{(m)*} \geq \gamma\right), \quad i=1,\dots,n'.
  \]
  \item[Sampled majority vote] If we choose to average the sampled binary responses \(Y^{(m)*}\) instead, we compute the predictions as
\[
  \hat Y_i = \mathbb I\left(\frac{1}{M} \sum_{m=1}^M Y_i^{(m)*} \geq \gamma\right), \quad i=1,\dots,n',
\]
where each \(Y_i^{(m)*}\) follows a Bernoulli distribution of parameter \(p_i^{(m)*}\), for \(m=1,\dots,M\).
\end{description}
Lastly, the variable selection method is essentially the same as in the case of linear regression: we select \(p\) times from each trajectory based on the posterior distribution of \(\tau\mid \mathcal D_n\), and then feed the reduced dataset to a multivariate binary classification procedure.

@@

As for the default values of the rest of hyperparameters in~\eqref{eq:prior-linear}, several comments can be made:
\begin{itemize}
  \item For the expected value \(b_0\) we propose to use (an approximation of) the MLE of \(\beta\) (see Section~\ref{sec:results} for details on the implementation). Our numerical studies showed that the results were much better with this choice than, say, with a random or null vector.
  \item We found that the parameter \(g\) does not have as much influence on the final result, and the experimentation suggests that \(g=5\) is a good value.
  \item Lastly, we observed that the choice of \(\eta\) can have a considerable impact on the final estimator. That is why, in an effort to normalize its scale, we consider a compound parameter \(\eta = \tilde \eta \lambda_{\max}(G_\tau)\), where \(\lambda_{\max}(G_\tau)\) is the largest eigenvalue of the matrix \(G_\tau\) and \(\tilde\eta > 0\) is the actual tuning parameter. This standardization technique has been used previously in the literature; see for example \citet{grollemund2019bayesian}.
\end{itemize}
